Transition path sampling based calculations of free energies of enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-05-05 tex/manuscript.tex
almost close to the finish line here

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories. The well known shooting algorithm begins with the choice of a 
random time slice from the biased reactive trajectory and perturbing the momenta of the system.
Then the dynamics is propagated for 250 steps forawards and backwards using $1\;fs$ time steps
resulting in a trajectory with a time period of $0.5\;ps$.
A total of 150 and 130 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively. 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The equilibrated structure of the MAT2A enzyme in complex with 
ATP, MET and two Mg$^{2+}$ ions is shown in Fig. \ref{fig:mat2a-equil}.
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 
MET along with 2 Mg$^{2+}$ ions.}
\label{fig:mat2a-equil}
\end{figure}
In the binding pocket the adenosyl moiety and the Phe250 and Ile117 residues 
are found to have $\pi-\pi$ interactions stabilizing the molecule. The triphosphate 
moiety forms hydrogen bonds with ... residues.

@@

Committor analysis and committor distribution analysis are used to deduce 
the transition states and the reaction coordinates, respectively. 
Committor analysis is carried out for a reactive trajectory 
from the TPS ensemble by choosing a time slice randomly from the trajectory, assigning
momenta picked at random from the Boltzmann distribution for all the atoms, and running 
dynamics for $250\;fs$. The commitment probability of a time slice is calculated by 
repeating this procedure several times and finding the fraction of trajectories that go 
to the reactant, product or neither. The transition state for the reaction is defined as 
the structure from the time slice that has equal probability to commit to the reactant 
and the product state. Repeating this process for several trajectories from the TPS
ensemble results in a collection of transition states called the separatrix. 
The reaction coordinate might contain not just the atoms from the molecules that 
are directly involved in the reaction but also residues from side chains that can 
promote the reaction. Taking this into account the committor distribution analysis
starts by constraining the residues and molecules that are theorized to be part of the 
reaction and using this constrained system trajectories are started

@@

The activation energy of this reaction was experimentally measured by 
Niland et al. \cite{Niland21Biochem60p791} where they measured the rate 
constant of the reaction at various temperatures while keeping the 
substrate concentrations near saturation. Using the Arrhenius equation 
they estimated the activation energy to be $17.27\;\pm\;1.5\;
\text{Kcal mol}^{-1}$. Compared to typical enzyme catalyzed 
reactions calculated activation energy is small.

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}
From our TPS based calculations we calculated the free energy barrier to be 
$16\;\text{Kcal mol}^{-1}$.
%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
The equilibrated structure for the ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$ cation is shown in Fig. \ref{fig:ada-equil}.
The Zn$^{2+}$ cation interacts with the His42, His44 and His226 residues
and stabilizes the OH$^{-}$ anion in the pocket. The N1 atom on the adensoyl 
molecule hydrogen bonds with the carboxylic acid hydrogen atom of the Glu229 
residue.  
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{./figures/ada-equil.png}
\caption{Equilibriated structure of ADA protein}
\label{fig:ada-equil}
\end{figure}
The bond forming ($dis(OC)$) and bond breaking ($dis(NC)$) distances 
for a reactive trajectory from the TPS ensemble is depicted 
in Fig. \ref{fig:ada-reactive-traj}.
\begin{figure}[ht!]
\includegraphics[scale=1.0]{figures/ada-diff60.pdf}
\caption{A reactive trajectory}
\label{fig:ada-reactive-traj}
\end{figure}
Committor analysis is performed to determine the transition state of this reaction
and the structure obtained from the calculations is shown in Fig. \ref{fig:ada-trans}.
At the transition state the N$1-$H bond distance is $1.1\;${\AA}, suggesting that the 
Glu229 residue has completed the proton donation to the adenosine molecule. This protonation
then triggers a base rotation which can be observed    
\begin{figure}
\centering
\includegraphics[scale=0.12]{blender-images/ada/new-ada-trans.png}
\label{fig:ada-trans}
\caption{The transition state structure of the reaction catalyzed by the ADA enzyme.
The formation of the Meissenheimer complex is the rate limiting step and the N-C distance is 
1.60 {\AA} and the O-C distance is 1.48 {\AA}.}
\end{figure}
Experimentally KIE studies 
from Schramm and coworkers \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
predict the transition state structures for the plasmodium falciparum, human and bovine 
ADA enzymes. They reported inhibition studies on the plasmodium vivax ADA by using 
MT coformycin as the inhibitor molecule.

The formation of Meissenheimer complex is part of the 
transition state as well as the partial protonation of the 
N1 atom on the adenosyl moity as observed from experiments. 
The reaction free energy is supposed to have two peaks with 
an intermediate state and we observe the same from our calculations.

\begin{figure}[ht!]
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\label{fig:ada-fenergy}
\caption{The free energy in $\text{Kcal}\;\text{mol}^{-1}$ as a function of the reaction 
parameter for the ADA catalyzed reaction. The Meissenheimer complex where the OH$^{-}$ anion
is bonded to the C atom and the leaving NH$_2$ moeity is also still bonded.}
\end{figure}

%------------------------------------------------------------------------------ 
\section{Conclusions}
%------------------------------------------------------------------------------ 
We have discussed the application of a scheme to calculate the 
free energies of enzymatic reactions within the TPS method using the MAT2A
and the ADA enzymes as examples. The MAT2A enzymatic reaction consists of 
the residues that are ... as part of its reaction coordinate. The reaction 
proceeds through a SN2 mechanism and the transition state consists of 
$dis(SC)$ being 2.3 and $dis(OC)$ being 2.03 {\AA}. From the calculations 
we obtain a free energy barrier of 16.0 kcal mol which is in good agreement 
with the 17.1 kcal mol measured from the experiments using the Arrhenius equation.
Furthermore we looked at the reaction catalyzed by the ADA enzyme and 
calculated the free energy profile for this reaction. Here we found two free energy
barriers separated by an intermediate state which is consistent with 
experimental findings. We calculate a free energy barrier of 25 kcal mol.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "Acknowledgement" section can be given in all manuscript
%% classes.  This should be given within the "acknowledgement"
%% environment, which will make the correct section or running title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgement}

>>>>

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories. The well known shooting algorithm begins with the choice of a 
random time slice from the biased reactive trajectory and perturbing the momenta of the system.
Then the dynamics is propagated for 250 steps forawards and backwards using $1\;fs$ time steps
resulting in a trajectory with a time period of $0.5\;ps$.
A total of 150 and 130 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively. 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value which is divided into 20 overlapping 
windows ($0.045$ {\AA} between neighboring windows). A reactive 
trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The equilibrated structure of the MAT2A enzyme in complex with 
ATP, MET and two Mg$^{2+}$ ions is shown in Fig. \ref{fig:mat2a-equil}.
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 
MET along with 2 Mg$^{2+}$ ions.}
\label{fig:mat2a-equil}
\end{figure}
In the binding pocket the adenosyl moiety and the Phe250 and Ile117 residues 
are found to have $\pi-\pi$ interactions stabilizing the molecule. The triphosphate 
moiety forms hydrogen bonds with ... residues.

@@

Committor analysis and committor distribution analysis are used to deduce 
the transition states and the reaction coordinates, respectively. 
Committor analysis is carried out for a reactive trajectory 
from the TPS ensemble by choosing a time slice randomly from the trajectory, assigning
momenta picked at random from the Boltzmann distribution for all the atoms, and running 
dynamics for $250\;fs$. The commitment probability of a time slice is calculated by 
repeating this procedure several times and finding the fraction of trajectories that go 
to the reactant, product or neither. The transition state for the reaction is defined as 
the structure from the time slice that has equal probability to commit to the reactant 
and the product state. Repeating this process for several trajectories from the TPS
ensemble results in a collection of transition states called the separatrix. 
The reaction coordinate might contain not just the atoms from the molecules that 
are directly involved in the reaction but also residues from side chains that can 
promote the reaction. Taking this into account the committor distribution analysis
starts by constraining the residues and molecules that are hypothesized to be part of the 
reaction coordinate and using this constrained system as the starting point 
trajectories are launched assigning momenta at random from the Boltzmann distribution at 
$300\;\text{K}$.

@@

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}
From our TPS based calculations we calculated the free energy barrier to be 
$16\;\text{Kcal mol}^{-1}$.
The activation energy of this reaction was experimentally measured by 
Niland et al. \cite{Niland21Biochem60p791} where they measured the rate 
constant of the reaction at various temperatures while keeping the 
substrate concentrations near saturation. Using the Arrhenius equation 
they estimated the activation energy to be $17.27\;\pm\;1.5\;
\text{Kcal mol}^{-1}$. Compared to typical enzyme catalyzed 
reactions calculated activation energy is small.  
\subsection{Adenosine deaminase reaction coordinates and free energy}
The equilibrated structure for the ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$ cation is shown in Fig. \ref{fig:ada-equil}.
The Zn$^{2+}$ cation interacts with the His42, His44 and His226 residues
and stabilizes the OH$^{-}$ anion in the pocket. The N1 atom on the adensoyl 
molecule hydrogen bonds with the carboxylic acid hydrogen atom of the Glu229 
residue.  
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{./figures/ada-equil.png}
\caption{Equilibriated structure of ADA protein}
\label{fig:ada-equil}
\end{figure}
The bond forming ($dis(OC)$) and bond breaking ($dis(NC)$) distances 
for a reactive trajectory from the TPS ensemble is depicted 
in Fig. \ref{fig:ada-reactive-traj}.
\begin{figure}[ht!]
\includegraphics[scale=1.0]{figures/ada-diff60.pdf}
\caption{A reactive trajectory}
\label{fig:ada-reactive-traj}
\end{figure}
The deamination is an aromatic nucleophilic substitution reaction 
where intermediate Meisenheimer complex is formed at the transition state
as shown in Fig. \ref{fig:ada-trans}. 
The approaching O atom of the hydroxide anion is at a distance of $1.48$ {\AA}
from the C6 atom whereas the N atom of the amine leaving group is at $1.60$ {\AA}.
The N$1-$H bond distance is $1.1\;${\AA} 
suggesting that the proton transfer from the Glu229 residue is almost complete.
\begin{figure}
\centering
\includegraphics[scale=0.12]{blender-images/ada/new-ada-trans.png}
\caption{The transition state structure of the reaction catalyzed by the ADA enzyme.
The formation of the Meissenheimer complex is the rate limiting step and the N-C distance is 
1.60 {\AA} and the O-C distance is 1.48 {\AA}.}
\label{fig:ada-trans}
\end{figure}
These observations are consistent with the 
experimental KIE studies from Luo et al. \cite{Luo07JAmChemSoc129p8008}
who predict the transition state structure for the deamination reaction 
catalyzed by the plasmodium falciparum ADA enzyme which shares
$72\%$ identity with the $pv$ADA enzyme that we study here. The frequency of the 
configurations (order parameter values) that are calculated within each of the 
windows is used to  
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\caption{The free energy in $\text{Kcal}\;\text{mol}^{-1}$ as a function of the reaction 
parameter for the ADA catalyzed reaction. The Meissenheimer complex where the OH$^{-}$ anion
is bonded to the C atom and the leaving NH$_2$ moeity is also still bonded.}
\label{fig:ada-fenergy}
\end{figure}
The free energy profile that we calculated is shown in Fig. \ref{fig:ada-fenergy}.
The free energy barrier is calculated to be $23\;\text{Kcal}\;\text{mol}^{-1}$.
\section{Conclusions}
In this study, we presented an application of the TPS method for the calculation 
of free energy profiles of enzymatic reactions. The order parameter along which the 
free energy profile is to be obtained is partitioned into windows similar to 
umbrella sampling. But instead of using bias potentials to sample the configurations
within the windows the shooting algorithm is used to
obtain an ensemble of trajectories according to the Monte carlo algorithm. A 
trajectory is accepted if it visits the window at least in one of the time steps. 
Using this ensemble of trajectories unbiased equilibrium probability distribution 
function of the order parameter of the reaction within each window is obtained.
Furthermore the free energy as a function of the order parameter is computed within 
each of the windows and the entirte profile is obtained by shifting the free energies
such that the function is continuous at the boundaries.

The MAT2A enzymatic reaction was analyzed in detail to elucidate the transition state,
reaction coordinate and the free energy profile. The reaction proceeds through a 
S$_\text{N}$2 mechanism, the d$_{\text{OC}}$ and d$_{\text{SC}}$ are $2.1$ {\AA}
and $2.3$ {\AA}, respectively. The reaction coordinate is determined to involve  
the Gln113, Ser114, Arg249 and Arg264 residues. 
The free energy barrier for this reaction was calculated to be 
$16.0\;\text{Kcal}\;\text{mol}^{-1}$ which is in good agreement 
with the experimentally measured activation energy of $17.27\;\pm\;1.5\;\text{Kcal mol}^{-1}$. 
Furthermore the reaction catalyzed by the $pv$ADA enzyme is analyzed in detail to obtain the  
transition state and calculate the free energy profile.
The transition state has a Meisenheimer complex like structure with the 
d$_\text{NC}$ and d$_\text{OC}$ being $1.60$ {\AA} and $1.48$ {\AA}, respectively.
The free energy barrier for this reaction for found to be $23\;\text{Kcal}\;\text{mol}^{-1}$
with a double well shape similar to that measured from experiments.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "Acknowledgement" section can be given in all manuscript
%% classes.  This should be given within the "acknowledgement"
%% environment, which will make the correct section or running title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgement}
